---
title: "What is the shape of data?"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Learning outcomes

By the end of this unit you will be able to:

- Visualize the frequency and impact of a single series of data

- Calculate the relative and cumulative relative frequency of the data for equally spaced class intervals of observations

- Graph the relative and cumulative relative frequency distributions

- Interpret the graphs and underlying calculations

## Visualize the data

It is often useful to organize a data sample into a frequency distribution. **Frequency** is simply the number or count of observations. This arrangement breaks up the data into classes or groups and shows the number of observations in each class. 

- The number of class intervals is usually an odd number between 5 and 15. Each interval is equally spaced. The intervals are not overlapping.

- We derive a relative frequency distribution by dividing the number of observations in each class by the total number of observations. 

- The sum of the frequencies must always equal the number of observations from the sample. The sum of the relative frequencies must always equal 1. 

There are three graphs we can make using this procedure:

1. A **histogram** is a bar graph of a frequency or relative frequency distribution, where classes are measured along the horizontal axis and frequencies or relative frequencies along the vertical axis. 

2. A **frequency polygon** is a line plot of a frequency distribution which joins the frequency of each class plotted at the class midpoint. A **relative frequency** polygon uses the relative frequency (fraction of observations in each class).

3. A **cumulative relative frequency distribution** shows, for each class, the total number of observations in
all classes up to and including that class. When plotted, this gives a cumulative distribution curve, or **ogive**.

## Build a histogram



## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

