---
title: "Yet Another FINAL -- BUAN 227 -- Spring 2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(digits = 4, scipen = 9999)
library(visNetwork)
library(tidyverse)
library(kableExtra)
library(ggplot2)
library(ggthemes)
#
name_Y0 <- "ungarbled price (Rs./Kg)"
name_Y <- "garbled price (Rs./Kg)"
name_X <- "quantity supplied"
description <- "Cochin peppercorns"
n <- 6
m_x <- 100
s_x <- 25
s_e <- 10
set.seed(1012)
X <- abs(rnorm(n, m_x, s_x))
Y <- abs(367 - 0.08 * X + rnorm(n, sd = s_e))
Y_0 <- abs(347 - 0.06 * X + rnorm(n, sd = s_e*1.10))
XY_df <- data.frame(X = X, Y = Y)
fit <- lm(Y ~ X, data = XY_df)
XY_df$predicted <- predict(fit)
XY_df$residuals <- residuals(fit)
e <- residuals(fit)
#
sumX <- sum(X)
sumY <- sum(Y)
sumY0 <- sum(Y_0)
sumXY <- sum(X*Y)
sumX2 <- sum(X^2)
Xbar <- mean(X)
Ybar <- mean(Y)
Y0bar <- mean(Y_0)
Xsd <- sd(X)
Ysd <- sd(Y)
Y0sd <- sd(Y_0)
diffbar <- Y0bar - Ybar
n0 <- n
diffsd <- sqrt(Y0bar^2/n0 + Ysd^2/n)
n0 <- n
N <- n
k <- 2
df <- N-k
b1 <- (N*sumXY - sumX*sumY) / (N*sumX2 - sumX*sumX)
b0 <- Ybar - b1*Xbar
sumY0dev2 <- sum((Y_0-Y0bar)^2)
sumYdev2 <- sum((Y-Ybar)^2)
sumXdev2 <- sum((X-Xbar)^2)
SST <- sum((Y-Ybar)^2)
SSE <- sum(e^2)
s_e <- sqrt(SSE / df)
s_b0 <- sqrt(s_e^2*(1/N + Xbar^2/sumXdev2))
s_b1 <- sqrt(s_e^2 / sumXdev2)
Rsq <- 1 - SSE / SST
t_b0 <- b0 / s_b0
t_b1 <- b1 / s_b1
alpha <- 0.05
tstar <- abs(qt(alpha / 2, df))
L_b0 <- b0 - tstar*s_b0
L_b1 <- b1 - tstar*s_b1
U_b0 <- b0 + tstar*s_b0
U_b1 <- b1 + tstar*s_b1
y_intercept_comment <- ifelse(b0 >= 0 , "positive", "negative")
slope_comment <- ifelse(b1 > 0, "positive", "negative")
hypo_b0 <- ifelse(-tstar < t_b0 , ifelse( t_b0 < tstar, "accept the null hypothesis", "reject the null hypothesis"), "reject the null hypothesis")
hypo_b1 <- ifelse(-tstar < t_b1 , ifelse( t_b1 < tstar, "accept the null hypothesis", "reject the null hypothesis"), "reject the null hypothesis")
t_diff <- (diffbar - 0) / diffsd
tstar_diff <- abs(qt(alpha / 2, n0 + n -2))
hypo_diff <- ifelse(-tstar_diff < t_diff , ifelse( t_diff < tstar_diff, "accept the null hypothesis", "reject the null hypothesis"), "reject the null hypothesis")
p_cum <- pt(t_diff, n0+n-2)
p_diff <- ifelse(t_diff > 0, 1 - p_cum, p_cum)
```

## A problem

Here is data for the `r name_Y0`, `r name_Y` and `r name_X` of `r description`. 

```{r}
#
header <- c(name_Y0, name_Y, name_X)
margin <- 1:N
col_0 <- Y_0
col_1 <- Y
col_2 <- X
table <- cbind(col_0, col_1, col_2)
colnames(table) <- header
rownames(table) <- margin
table %>% 
  kable(digits = 2)
```

[Malabar pepper](https://en.wikipedia.org/wiki/Malabar_pepper) is classified under two grades known as garbled and un-garbled. The garbled variety (higher grade) is black in color with a wrinkled surface. The ungarbled variety has a wrinkled surface and the colour varies from dark brown to black.

Here are calculations for the `r name_Y` and `r name_X` data:

```{r}
#
header <- c("estimate", "standard deviation")
margin <- c("intercept", "slope")
row_1 <- c(b0, s_b0)
row_2 <- c(b1, s_b1)
table <- rbind(row_1, row_2)
colnames(table) <- header
rownames(table) <- margin
table %>% 
  kable(digits = 4) 
#%>%
#    kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

$N= `r N`$, $\Sigma_{i=1}^N Y_i= `r sumY`$, $\Sigma_{i=1}^N X_i= `r sumX`$, $\Sigma_{i=1}^N X_iY_i= `r sumXY`$, $\Sigma_{i=1}^N X_i^2= `r sumX2`$, $\Sigma_{i=1}^N (Y_i - \bar Y)^2= `r SST`$, $\Sigma_{i=1}^N (Y_i - \bar Y)^2= `r sumYdev2`$, $\Sigma_{i=1}^N (X_i - \bar X)^2= `r sumXdev2`$, $\Sigma_{i=1}^N e_i^2= `r SSE`$, $s_e= `r s_e`$, $R^2=`r Rsq`$, $\alpha = `r alpha`$ with $|t| = `r tstar`$ for appropriate degrees of freedom. 

For the `r name_Y0` data here are additional calculations where $Y =$ `r name_Y0`:

$\Sigma_{i=1}^N Y_i= `r sumY0`$, $\Sigma_{i=1}^N (Y_i - \bar Y)^2= `r sumY0dev2`$

For complete credit you must draw pictures, show formulas, show all calculations needed, interpret process and results.

## Question 1: Should we use two prices?

We are going to try to forecast prices. There are two prices at Cochin. Do they act as if one world price? 

a. Formulate an hypothesis to test a claim that there is one world price. 

b. Test the hypothesis and interpret the results.

## Question 2: What is the model?

Assume a linear relationship between the independent variable `r name_X` and dependent variable `r name_Y`. Include in this description the following with interpretations:

a. A scatterplot that depicts the relationship between dependent and independent variables. A straight line through the data, consistent with the calculations, that depicts deviations of scatter points from the line.

b. Validate the calculations of the sample mean and standard deviation of the intercept parameter as well as the percentage of variation in the dependent variable explained by the model.

## Question 3: How confident are we about the estimates?

Given a level of significance of `r alpha*100`\%, 

a. Draw an appropriate probability distribution to construct an estimate of the confidence interval for $b_0$ and $b_1$. Be sure to include the upper and lower bounds on the distribution in terms of the number of standard deviations from the mean zero of this distribution.

b. Calculate the upper and lower bounds for the confidence intervals of $b_0$ and $b_1$. Interpret the results.

## Question 4: How meaningful are the estimates?

a. Formulate testable hypotheses of the meaningfulness of the mean slope and Y-intercept of our model. Interpret.

b. Test the hypotheses and interpret.
