---
title: "Probability Distributions: Poisson"
author: "Copyright 2016. William G. Foote."
date: "`r Sys.Date()`"
fontsize: 10pt
theme: "Madrid"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
knitr::opts_chunk$set(tidy = TRUE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=36))
knitr::opts_chunk$set(size = "small")
knitr::opts_hooks$set(fig.width = function(options) {
  if (options$fig.width < options$fig.height) {
    options$fig.width = options$fig.height
  }
  options
})
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```

# Imagine this
- Your corporate website get 2 hits per minute
- There have been 3 pandemics in the last 100 years
- Your satellite TV fails to record a show at least 4 times a month
- Your most important client has 6 meltdowns a month

## What do you want to know?
How probable any and / or all of the events will happen!

## Why?
So you can plan your life a bit better -- allocation scarce time and money more efficiently, at the very least

# Enter M. Poisson
## Key assumptions
The Poisson random variable satisfies the following conditions:

1. The number of successes in two disjoint (non-overlapping) time intervals is independent.
2. The probability of a success during a small time interval is proportional to the entire length of the time interval.


Apart from disjoint time intervals, the Poisson random variable also applies to disjoint regions of any kind of space, such as geographical and market.

# Applications

- The number of deaths by horse kicking in the Prussian army (first application)
- Birth defects and genetic mutations
- Rare diseases
- Car accidents
- Traffic flow and ideal vehicle gap distance
- Number of typing errors on a page
- Hairs found in McDonald's hamburgers
- Extinction of an endangered animal in Africa
- Failure of a machine in one month
- Frequency of cyber attacks in a year

## Use the binomial when
you need to find the probability of the number of successes in _so many trials_.

## Use the Poisson distribution when
you need to find the probability of a number of successes _unit of time or space_.

# Let's calculate

- First define the number of success as $X$. 
- Next, suppose successes occur with intensity (same as: at a frequency or rate) of $\lambda$ per unit time or space. 
- Then, the probability of a specified number of successes $X = x$ per unit time (or space) is
\[
P(X = x) = \frac{\lambda^{x}e^{-x}}{x!}
\]
where $e = 2.71828$ is the base of the natural logarithm.

# Try this
We estimate that 6 accidents ("vehicular contacts") per month occur at the intersection of Crosby and Westchester Avenues in the Bronx. 

1. What is the probability of 3 accidents in any month?
2. Plot the distribution of accidents from 0 to 20.
3. what is the probability of 3 accidents or less in any month?

# Results
1. What is the probability of 3 accidents in any month? We let $X$ be the number of possible accidents per month at this intersection. Out data is $X = x = 3$ and $\lambda = 6$. We can then  compute
\[
\begin{array}{c}
P(X = 3) = \frac{\lambda^{x}e^{-x}}{x!} \\ \\
= \frac{6^{3}e^{-3}}{3!} \\ \\
= \frac{(216)(0.002478)}{3x2x1} \\ \\
= \frac{0.5354105}{6} \\ \\
= 0.08923508
\end{array}
\]
that is, 8.9\%.

***
2. Plot the distribution of accidents from 0 to 20.

```{r, echo = FALSE}
gpois = function(mu, a=NA,b=NA,calcProb=(!is.na(a) | !is.na(b)),quantile=NA,calcQuant=!is.na(quantile),title=NA)
{
  sd = sqrt(mu)
  low = max(0, round(mu - 3 * sd))
  high = round(mu + 5 * sd)
  values = low:high
  probs = dpois(values, mu)
  plot(c(low,high), c(0,max(probs)), type = "n", xlab = "Number of Occurrences",
      ylab = "Probability",
      main = paste(title))
  lines(values, probs, type = "h", col = 2)
  abline(h=0,col=3)
  if(calcProb) {
    if(is.na(a)){ a = 0 }
    if(is.na(b)){
     a = round(a)
     prob = 1-ppois(a-1,mu)
     sub(paste("P(",a," <= X <= ",b,") = ",round(prob,6),sep="") ,line=0,col.main=4)
     u = seq(max(c(a,low)),high,by=1)
    }
    else {
     if(a > b) {d = a; a = b; b = d;}
     a = round(a); b = round(b)
     prob = ppois(b,mu) - ppois(a-1,mu)
     title(paste("P(",a," <= X <= ",b,") = ",round(prob,6),sep="") ,line=0,col.main=4)
     u = seq(max(c(a,low)),min(c(b,high)),by=1)
    }
    v = dpois(u,mu)
    lines(u,v,type="h",col=4)
  }
  else if(calcQuant==T) {
    if(quantile < 0 || quantile > 1)
      stop("quantile must be between 0 and 1")
    x = qpois(quantile,mu)
    title(paste("The ",quantile," quantile = ",x,sep=""),line=0,col.main=4)
    u = 0:x
    v = dpois(u,mu)
    lines(u,v,type="h",col=4)
  }
  return(invisible())
}
gpois(6, a=0, b = 3, title="Accidents Will Happen")
```

***
3. What is the probability of 3 accidents or less in one month?

```{r, echo = FALSE, eval = FALSE}
dpois(0:3,6)
```

# Short exercises

1. Customers pass through the doors of a retail outlet at an average rate of 300 per hour. 

- Find the probability that none passes in a given minute.
- What is the expected number passing in two minutes?
- Find the probability that this expected number actually pass through in a given two-minute period.

_Answers_ : 0.0006, 10, 12.5\%

2. A company makes windmill rotors. The probability a rotor is defective is 0.01. What is the probability that a sample of 300 rotors will contain exactly 5 defective rotors? _[Use  both the Poisson and the Binomial distributions to arrive at an answer.]_

_Answer_ : a little over 10\%

# Not so short exercises

The number of calls into your call center averages 3 per minute.

(a) Find the probability that no calls come in a given 1 minute period. _Answer_: 4.978\%
(b) Assume that the number of calls arriving in two different minutes
are independent. Find the probability that at least two calls
will arrive in a given two minute period. _Answer_: 98.26\%

# Answers
Let $X_1$ and $X_2$ be the number of calls coming in the first and second
minutes respectively. We want $P(X_1 + X_2 \geq 2)$.

1. Let's organize the answer this way
\[
P(X_1 + X_2 \geq 2) = 1-P(X_1 + X_2 < 2)
\]
In this way we only need to look at no calls (0 minutes) and 1 minute calls, or, for each possible sequence of 0 and 1 minute calls each sequence happening together (intersection) and independently. We have
$1-P(X_1 + X_2 < 2) = 1-(P(X_1 = 0 \cap X_2 = 0)$
$+ P(X_1 = 1 \cap X_2 = 0) + P(X_1 = 0 \cap X_2 = 1))$
which, in turn, equals
$= 1 - [P(X_1 = 0) P(X_2 = 0)$
$+ P(X_1 = 1) P(X_2 = 0) + P(X_1 = 0) P(X_2 = 1)]$

***
2. We already know from the first answer that
\[
P(X = 0) = \frac{e^{-3}3^0}{0!} = e^{-3} 
\]
which equals 0.0498. We can then compute
\[
P(X = 1) = \frac{e^{-3}3^1}{1!} = 3exp(-3)
\]
which equals 0.1494.

***
3. Putting this altogether we have
$P(X_1 + X_2 \geq 2) =$
$1- [(0.0498)(0.0498)+(0.1494)(0.0498)+(0.0498)(0.1494)]$
$= 1-0.01736$
$= 0.9826$

or 98.26\%.