---
title: "Locating to the Bronx?"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggmap)
library(stringr)
library(viridis)
library(plotly)
#
api_key <- "AIzaSyA5XmoJgeHOcNNxqixIQojokXOB_7YwOMY"
register_google(api_key, write = TRUE)
```

# Situation, complications, and a question

Imagine that a business wants to locate its offices in the Bronx to take advantage of NYC incentives and relatively low rents. The operation is 24/7, so some workers will be coming to and going from work late at night. Are there any areas so unsafe that the business should avoid these neighborhoods?

# Load and inspect data

Open NYC Data has NYPD complaint data for the past few months at <https://data.cityofnewyork.us/Public-Safety/NYPD-Complaint-Data-Current-YTD/5uac-w243/data>, We will use `tidyverse` to read and inspect the data structurally and visually. A simple `lat` and `lon` (latitude and longitude) scatter plot will indicate light and dark areas for us to inspect the data.

```{r}
library(tidyverse)
library(ggmap)
library(stringr)
library(viridis)
library(plotly)

#Load data

bx_crime <- read_csv("bx-crime.csv")

# Inspect data

bx_crime %>% glimpse()
ggplot() +
  geom_point(data = bx_crime, aes(x = lon, y = lat), alpha = .05)
```

# Get map

we need a satellite map of the Bronx and will use `ggmap` to download and view the map. But first you need to get a Google geocoding API key. As of mid-2018, the Google Maps Platform requires a registered API key. While this alleviates previous burdens (e.g. query limits), it creates some challenges as well. You will have to setup an account with Google, enable the relevant APIs, and then tell R about the user's setup.

To obtain an API key and enable services, [go to this site.](https://cloud.google.com/maps-platform/). 

To tell ggmap about your API key, use the register_google() function, e.g. register_google(key = "mQkzTpiaLYjPqXQBotesgif3EfGL2dbrNVOrogg", write = TRUE) (that's a fake key). This will set it permanently where it is stored in your .Renviron file, and that will be accessed by ggmap persistently across sessions.

```{r}
map_bx <- get_map("Bronx", zoom = 12, maptype = 'satellite')
ggmap(map_bx)
```

# Build heat map

Again using `ggmap`, based on `ggplot2`, we use the `stat_density2d` function to view an overlay of the incidents scatterplot onto the Google satellite map. Titles, labels, density colors (courtesy of `viridis` and the `inferno` color option), along with theme formatting complete the picture.

```{r}
p <- ggmap(map_bx) +
  stat_density2d(data = bx_crime, aes(x = lon, y = lat, 
      fill = ..density..), geom = "tile", contour = F, alpha = .5) +
  scale_fill_viridis(option = "inferno") +
  labs(title = str_c("BX has largest concentration of crime\n"
                     ,"near Fordham Rd., Morrisania, and the South Bronx"
                     ),
       subtitle = str_c("2017-18 source:", "\nhttps://data.cityofnewyork.us/Public-Safety/NYPD-Complaint-Data-Current-YTD/5uac-w243/data"),
       fill = str_c("Number of", "\ncrime incidents")
       ) +
  theme(text = element_text(color = "#444444"),
        plot.title = element_text(size = 13, face = 'bold'),
        plot.subtitle = element_text(size = 8),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank()
        ) +
  guides(fill = guide_legend(override.aes= list(alpha = 1)))
p
```

# Decision?

Not quite yet! There are neighborhoods to assess further. Here we have several neighborhoods suggested by the clusters of high and low density incidents to ponder. Our next stop might be a visit to the Economic Development Corporation for data about economic and commercial zones in the Bronx.
