---
title: "DriveTime Used Cars"
output: 
  html_document:
    toc: true
    toc_float: true
    source_code: embed
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = TRUE)
library(tidyverse)
library(plotly)
library(GGally)
library(knitr)
library(kableExtra)
#
# function to specify scatterplot smoothers in GGally::ggpairs()
#
smooth_over <- function(data, mapping, method="loess", ...){
  # function to specify scatterplot smoothers in GGally::ggpairs()
  # try this with ggpairs(data, lower = list(continuous = smooth_over)) for loess
  # try this with ggpairs(data, lower = list(continuous = wrap(smooth_over, method="lm"))) for lm
      p <- ggplot(data = data, mapping = mapping) + 
      geom_point() + 
      geom_smooth(method=method, ...)
      p
    }
```

## Driving in my car

The [DriveTime](https://www.drivetime.com/used-cars) has current market data for used cars across the United States. This table summarizes miles driven and prices paid for Hyundai Elantras in California.


```{r cars-data}
cars_data <- read_csv("drive_time_sedans.csv")
cars_hyundai_elantra_CA_summary <- cars_data %>% select(total.cost, mileage, makex, make.model, state) %>% filter(state == "CA" & makex == "HYUNDAI" & make.model == "HYUNDAI.ELANTRA") %>% select(price = total.cost, miles = mileage) %>% gather(key = "key", value = "value") %>% group_by(key) %>% summarize(min = min(value), q1 = quantile(value, 0.25), median = quantile(value, 0.50), q3 = quantile(value, 0.75), max = max(value), mean = mean(value), sd = sd(value), mad = sum(abs(value - mean(value))/length(value))) %>% mutate(IQR = q3 - q1, cv = sd / mean)
car_tbl <- cars_hyundai_elantra_CA_summary
car_tbl %>% 
  kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>% 
  column_spec(1, bold = T, color = "blue", border_right = FALSE)
```

A Tukey (1977) box plot visualized the table.

```{r cars_boxplot}
cars_hyundai_elantra_CA <- cars_data %>% select(total.cost, mileage, makex, make.model, state) %>% filter(state == "CA" & makex == "HYUNDAI" & make.model == "HYUNDAI.ELANTRA") %>% select(miles = mileage, price = total.cost)
p <- cars_hyundai_elantra_CA %>% 
    ggplot(aes(y = price)) +
     geom_boxplot()
ggplotly(p)
p <- cars_hyundai_elantra_CA %>% 
    ggplot(aes(y = miles)) +
     geom_boxplot()
ggplotly(p)
ggpairs(cars_hyundai_elantra_CA, lower = list(continuous = wrap(smooth_over, method="lm")))
```

There almost seem to be two clusters through the 60,000 mile mark. Let's check that out.

## The down-low

Let's divide the observations into two buckets: prices for miles < 60,000 and prices for miles $\geq$ 60000. We will summarize and view scatter matrices.

Here's the lower mileage cluster.

```{r cars-cluster-low}
cars_low_miles <- cars_hyundai_elantra_CA %>% 
  filter(miles < 60000)
ggpairs(cars_low_miles, lower = list(continuous = wrap(smooth_over, method="lm")), title = "Cars with mileage < 60000")
#
cars_low_miles_summary <- cars_low_miles %>% gather(key = "key", value = "value") %>% group_by(key) %>% summarize(min = min(value), q1 = quantile(value, 0.25), median = quantile(value, 0.50), q3 = quantile(value, 0.75), max = max(value), mean = mean(value), sd = sd(value), mad = sum(abs(value - mean(value))/length(value))) %>% mutate(IQR = q3 - q1, cv = sd / mean)
car_tbl <- cars_low_miles_summary
car_tbl %>% 
  kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>% 
  column_spec(1, bold = T, color = "blue", border_right = FALSE)
```

Relative to the overall sample, this filtered sample has a slightly flatter correlation and definitely a different set of miles and price distributions as evidenced in the summary table.

What about higher mileage cars?


```{r cars-cluster-high}
cars_high_miles <- cars_hyundai_elantra_CA %>% 
  filter(miles >= 60000)
ggpairs(cars_high_miles, lower = list(continuous = wrap(smooth_over, method="lm")), title = "Cars with mileage >= 60000")
#
cars_high_miles_summary <- cars_high_miles %>% gather(key = "key", value = "value") %>% group_by(key) %>% summarize(min = min(value), q1 = quantile(value, 0.25), median = quantile(value, 0.50), q3 = quantile(value, 0.75), max = max(value), mean = mean(value), sd = sd(value), mad = sum(abs(value - mean(value))/length(value)),) %>% mutate(IQR = q3 - q1, cv = sd / mean)
car_tbl <- cars_high_miles_summary
car_tbl %>% 
  kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>% 
  column_spec(1, bold = T, color = "blue", border_right = FALSE)
```

Again, these ranges are very different than the whole sample, and even less correlated than the low mileage cars.

