---
title: "Midterm Assessment -- BUAN 227"
author: "Fall 2018"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, warning = FALSE, message = TRUE)
options(digits = 5, scipen = 99999)
library(knitr)
library(magrittr)
library(kableExtra)
```

Show all of your work. Draw pictures. Make a point; justify it; illustrate it with an example. 

A utility company is concerned about the ability of its vehicle operators to drive safely in its service region. It decides to sample a population of accidents in the 10461 zip code. One particular intersection comes to light. Analysts estimate that 2 accidents ("vehicular contacts") per month occur at the intersection of Crosby and Westchester Avenues in the 10461 zip code in the Bronx. 

Here are some preiminary calculations.

```{r}
data <- read.csv("data/ex-table.csv", stringsAsFactors = F)
colnames(data) <- c("Index",	"Accidents",	"Probability",	"Deviations Squared",	"Deviations Cubed", "Quartic Deviations")
data %>% 
  kable() #%>%
    #kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

Based on these calculations here are some metrics.

```{r}
data <- read.csv("data/ex-metrics.csv", stringsAsFactors = F)
colnames(data) <- c("Metric",	"Level")
data %>% 
  kable() #%>%
    #kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```



1. Is there a random variable in this analysis? What kind of data do we have here?

2. Calculate and interpret the expected value of accidents. Describe and interpret the standard deviation, skewness, and kurtosis. 

3. What is the probability that there are either between 0 and 1 accidents (inclusive) or between 4 and 5 accidents? What is the probability that there are between 2 and 3 accidents (inclusive)? Explain.

4. Calculate and interpret the z-score of 5 accidents. According to the Chebyshev rule what is the probability of finding accidents that many standard deviations from the expected value? Does this check out with the data?

5. Discuss the pros and cons of 3 different scale (width, volatility, dispersion) metrics for a distribution of data.