---
title: "The many faces of data"
output: 
  html_document:
    toc: true
    toc_float: true
---
<script>
function showText(y) {
    var x = document.getElementById(y);
    if (x.style.display === "none") {
        x.style.display = "block";
    } else {
        x.style.display = "none";
    }
}
</script>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)



suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(visNetwork))
```

<script>
function showText(y) {
    var x = document.getElementById(y);
    if (x.style.display === "none") {
        x.style.display = "block";
    } else {
        x.style.display = "none";
    }
}
</script>

## Learning outomes

1. Identify the different types of data

2. Describe the use of different types of data to explore answers to business questions

3. Identify a research question and justify the the use of different types of data

Data is all around us. It is first an observation of an event like number of store visits, and later might be observations of the result of a conceptual framework like customer satisfaction. Where we make a store transaction would result in latitude and longitude data. There is also data that describes other data like kilo-watt-hours as a unit of measurement for electricity usage.

## March on

Here is a classic example of the many types of data we will confront.

[Minard: Napoleon's march to and from Moskow](poster_OrigMinard.gif)

How many data streams can you count in this graphic?

## Air pollution and birth outcomes

Businesses will often attempt to measure the impact of their operations on the populations they serve or are located near. Energy companies manufacture electricity, refined oil products, downstream basic and derived chemicals, and create polluting substances in the process. A California-based energy company wanted to gauge the impact of air pollution on births. Visit this [EPA site for more information on the management of air quality](https://www.epa.gov/air-quality-management-process/managing-air-quality-human-health-environmental-and-economic).

Researchers collected data to examine the relationship between air pollutants and preterm births in Southern California. During the study air pollution levels were measured by air quality monitoring stations. Specifically, levels of carbon monoxide were recorded in parts per million, nitrogen dioxide and ozone in parts per hundred million, and coarse particulate matter (PM10) in $\mu/m^3$. Length of gestation data were collected on 143,196 births between the years 1989 and 1993, and air pollution exposure during gestation was calculated for each birth. The analysis suggested that increased ambient PM10 and, to a lesser degree, CO concentrations may be associated with the occurrence of preterm births.

In this study, identify

- the cases,

- the variables and their types, and

- the main research question.

### Classifying data types

A familiar example of a hierarchy is the organizational chart. This chart visualizes the reporting structure of an organization.

```{r}
visNetwork(main = list(text = "Hierarchy Example: Student Life",
 style = "font-family:Comic Sans MS;color:#ff0000;font-size:15px;text-align:center;"),
        dot = 'dinetwork{ node [shape=box];  
           "VP Student Life" -> "Residence Director" ; 
           "VP Student Life" -> "Program Director" ; 
           "Residence Director" -> "Residence Advisor-Alpha" ;
           "Residence Director" -> "Residence Advisor-Beta" ;
           "Residence Director" -> "Residence Advisor-Gamma" ;     
           }'
           ) %>%  visNodes(color = list(background = "white", border = "blue", 
    highlight = "yellow")) %>% visHierarchicalLayout(nodeSpacing = 200, sortMethod = "directed")
```

You can access this [article about hierarchy](https://en.wikipedia.org/wiki/Hierarchy) for more information.



1. What is a hierarchy?

</br>
<button onclick="showText('myDIV1')">show / hide</button>
<div id="myDIV1" style="display:none;">
</br>

A hierarchy is an analytical technique that takes a group of objects and asks two questions:

1. How are the objects (nodes) related to one another (edges)? (just a network)

2. What objects are parents (higher level) or children (lower level) of one another

Notice that the departments and their children nodes are mutually exclusive of one another.

</div>
</br>

In each department of science there are hierarchies and networks of metadata. Metadata are data about data. Statistics mirror life, and in life we find ourselves classifying the world around us to make some systematic sense of it.

Here is a potential data classification hierarchy we might use with this research example.

```{r}
visNetwork(main = list(text = "Data Classification",
 style = "font-family:Comic Sans MS;color:#ff0000;font-size:15px;text-align:center;"),
        dot = 'dinetwork{ node [shape=box];  
           "Data" -> "Numeric" ;
           "Data" -> "Categorical" ;
           "Numeric" -> "Continuous" ;
           "Numeric" -> "Discrete" ;
           "Categorical" -> "Nominal" ; 
           "Categorical" -> "Ordinal" ; 
           }'
           ) %>%  visNodes(color = list(background = "white", border = "blue", 
    highlight = "yellow")) %>% visHierarchicalLayout(nodeSpacing = 200, sortMethod = "directed")
```


2. Map the data in the research example to the hierarchy.

</br>
<button onclick="showText('myDIV2')">show / hide</button>
<div id="myDIV2" style="display:none;">
</br>

Let's classify the data in the the research example with this hierarchy.

- Cases: or "observations" include measurements relative to 143,196 births between the years 1989 and 1993 (Discrete - simply a convenient lexicographic index, not necessarily ordinal).

- Variables: air quality station (discrete - nominal), year (discrete - nominal) length of gestation (continuous - ratio - has a 0 starting basis), levels of carbon monoxide were recorded in parts per million (continuous - ratio), nitrogen dioxide and ozone in parts per hundred million (continuous - ratio), and coarse particulate matter (PM10) in $\mu/m^3$ (continuous - ratio). All continuous variables are based from zero to some positive number. There appear to be no continuous - interval measures that would allow negative numbers (e.g., temperature in degrees centigrade).

What further questions might be possibly related to these measures? Do these questions anticipate the collection of discrete- ordinal variables? For example, are there preferences for living in one region versus another (ordinal)? How satisfied are respondents with air quality?

</div>
</br>

3. Describe the usefulness of the components of your data classification hierarchy to answer the main research question above.

The data classification hierarchy is a theory to be tested. A **canon** is a general law, rule, principle, or criterion by which something is judged. Let's test this hierarchy against these canons of empirical method, expressed as questions for consideration (see [and astrophysicist's view of these canons](https://adh-sj.info/canons.php), both based on [Bernard Lonergan's _Insight_, Longmans, London 1957, chapter III](https://www.amazon.com/Insight-Understanding-Collected-Bernard-Lonergan/dp/0802034551) :

- **Relevance and Selection**: Does the hierarchy use terms and relationships that hint at attributes in the data in the example? Does the hierarchy anticipate, and possibly raise further questions about the data? Does the hierarchy select relevant data and de-select irrevelant data.

- **sufficiency**: If you use another node or relationship in the hierarchy, does that add anything new to the description that is relevant?

- **Parsimony**: is the hierarchical description of data itself verifiable in the data?

- **Completeness**: does this hierarchy account for all of the aspects you observe about data in this example? In other words, does the hierarchy somehow employ all of the data available, relevant, sufficient, and verifiable in this example.
</br>

<button onclick="showText('myDIV3')">show / hide</button>
<div id="myDIV3" style="display:none;">
</br>

By canon:

- **Relevance and Selection**: Does the hierarchy use terms and relationships that hint at attributes in the data in the example? Does the hierarchy anticipate, and possibly raise further questions about the data? Does the hierarchy select relevant data and de-select irrevelant data.

_Yes it does. Anticipated are variables that are non-integer (as well as integer) and simply unordered categorical data. Other types are possible, but not present._

- **Sufficiency**: If you use another node or relationship in the hierarchy, does that add anything new to the description that is relevant?

_No it does not. It seems that the continuous and discontinuous numeric variables are sufficient for numeric variables if only because they are mutually exclusive. The same for nominal (unordered) and ordinal (ordered). If fact the hierarchy **overfits** the data in this exercise by the inclusion of discrete and ordinal data types. Including these anticipate relevancy and selection in other studies._ 

- **Parsimony**: is the hierarchical description of data itself verifiable in the data?

_Yes it is. Again our data seems well matched to the lowest level of the hierarchy. Further levels might delineate attributes of the previous level. For example, add another component, say integer. But integer is just another way of saying discrete variable. so this doesn't add anything. Are there other categories of numeric or categorical data? There do not seem so._

- **Completeness**: does this hierarchy account for all of the aspects you observe about data in this example? In other words, does the hierarchy somehow employ all of the data available, relevant, sufficient, and verifiable in this example.

_Yes it does. The data in the example are continuous, discrete, nominal, and ordinal. There does not seem to be any other relevant data available within the field of this example._

</div>
</br>

## Something a bit more complicated

Electricity customers now have their choice of service providers. The service providers can offer fixed and flexible rates for electricity usage. Usage data can be collected using remote sensing technology. An example of usage data are the following text strings for a facility in New York City.

```{r usage}
usage <- data.frame(index = 1:6, usage = c(  "3 PM  Mon 24th-Mar-2014___0.384 kwhNA_R",  
"5AM  15-Aug-2014___1.201  kwh5_C",
"__8PM Thu 20-Mar-2014____1.523 kwh_4_C",  
"  6PM  23rd-Apr-2014___0.424  kwh_5_R",
"_1AM  Friday 19th-Dec-2014___0.209  kwh_3_R",  
"_5AM  Tue 19th-Aug-2014___1.228  kwh_4_C"
))
usage %>%
  kable("html", escape = F) %>% 
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```

All of the raw data is in a text string. We can parse the text into time of day, date, usage level, units of measurement, a quality rating (1-5), and a customer type ( _R_ for residential and _C_ for commercial) . Elements such as blank spaces and underscores (i.e., "_") would be eliminated as delimiters.

Does the data type hierarchy still apply?

</br>
<button onclick="showText('myDIV8')">show / hide</button>
<div id="myDIV1" style="display:none;">
</br>
Let's modify the hierarchy a bit to include the text string raw data.

```{r}
visNetwork(main = list(text = "Electricity Usage Data Classification",
 style = "font-family:Comic Sans MS;color:#ff0000;font-size:15px;text-align:center;"),
        dot = 'dinetwork{ node [shape=box];  
           "String: 3 PM  Mon 24th-Mar-2014___0.384 kwh3_R" -> "Numeric" ;
           "String: 3 PM  Mon 24th-Mar-2014___0.384 kwh3_R" -> "Categorical" ;
           "Numeric" -> "Continuous" ;
           "Continuous" -> "Ratio" ;
           "Ratio" -> "Usage: 0.384";
           "Numeric" -> "Discrete" ;
           "Discrete" -> "Time/Date: 3 PM  Mon 24th-Mar-2014";
           "Categorical" -> "Nominal" ;
           "Nominal" -> "Customer Type: R";
           "Nominal" -> "Units of Measurement: kwh";
           "Categorical" -> "Ordinal" ;
           "Ordinal" -> "Quality Rating: 3"
           }'
           ) %>%  visNodes(color = list(background = "white", border = "blue", 
    highlight = "yellow")) %>% visHierarchicalLayout(nodeSpacing = 200, sortMethod = "directed")
```
Even though the time "3 PM"  be would be arsed into 15:00 hours (on a 24-hour clock), the "PM" string could be used as discrete data to mark the second 12 hours of a day. Both usage and time are ratio types since both would start at a zero point. We might argue that days from 1/1/1900 to the date in the data is discrete as might be hours ending 3PM.

</div>

## What have we learned?

**Why does any of this matter?** The general question is bias and manipulation are possible if we fail to select all relevant, verifiable data that sufficiently describes and explains our research, decisions, and raises questions for further inquiry.

- Systematic inquiry uses structures to name, associate, analyze, and conclude the properties of data

- A proposed classification hierarchy anticipates the data used to analyze a pratical research problem

- A test of the hierarchy using rules of empirical methodology confirms the validity of the classifications of the data and reaches into further questions for inquiry

- A provisional structure helps us to identify and correct bias


## Try this

Apply the data hierarchy to the county health outcome rankings and data base for states in the US. Find the types of data and apply the canons of empirical method to your findings. Choose Other measures > socio-economic factors for your investigation. [You can navigate to this site for an interactive view.](https://www.countyhealthrankings.org/app/new-york/2019/measure/outcomes/1/data)