---
title: "The shape of data"
output: 
  html_document:
    toc: true
    toc_float: true
---

<script>
function showText(y) {
    var x = document.getElementById(y);
    if (x.style.display === "none") {
        x.style.display = "block";
    } else {
        x.style.display = "none";
    }
}
</script>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)



suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(stringr))
suppressPackageStartupMessages(library(visNetwork))
```

<script>
function showText(y) {
    var x = document.getElementById(y);
    if (x.style.display === "none") {
        x.style.display = "block";
    } else {
        x.style.display = "none";
    }
}
</script>

## Learning outomes

In this unit we will:

1. Arrange numeric data into groups (also known as bins, classes, intervals)

2. Count the number of data points in each group: the frequency

3. Transform the absolute frequency (count) into a relative frequency and cumulative relative frequency

4. Plot on the vertical axis the relative and cumulative relative frequencies of the horizontal axis midpoints of each bin

5. Use the plots to describe the shape of the data and the implications of shape for decisions

For the plots we will use a special bar graph called a histogram. We could also connect the midpoints with a line to produce a polygon graph. These graphs will help us answer two questions:

1. What is the range of impact of the variable?

2. How often do values and ranges of values of the variable occur?


## How well did we do?

Imagine we run a distribution center for a major appliance manufacturer. Key indicators of our operational performance include the number of on-time in-full orders delivered, time from receipt of order to delivery, returns, and overall service level. Time, cost, and quality are the hallmarks of a well run supply chain. 

Here is a sample of 20 separate on-time in-full orders for the past month. Each observation is the number of items in the order.

```{r orders}
orders <- sample(x = 76:100, size=20, replace=TRUE, prob=c(rep(0.10, 5), rep(0.30, 8), rep(0.40, 7), rep(0.15, 4), rep(0.20, 1)))
orders_tbl <- orders[order(orders)]
column_names <- c(1:length(orders))
row_names <- "orders"
table_data <- matrix(orders, nrow = 1, ncol = length(orders), byrow = TRUE) 
colnames(table_data) <- column_names
rownames(table_data) <- row_names
table_data %>% 
  kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>% 
  column_spec(1, bold = T, color = "red", border_right = TRUE) 
```

The data is a mix of various values with an index in the top row.

Our procedure straitforward:

1. Arrange the data from lowest to highest values

```{r sort}
orders_sort <- orders[order(orders)]
column_names <- c(1:length(orders))
row_names <- "sorted \n orders"
table_data <- matrix(orders_sort, nrow = 1, ncol = length(orders_sort), byrow = TRUE) 
colnames(table_data) <- column_names
rownames(table_data) <- row_names
table_data %>% 
  kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>% 
  column_spec(1, bold = T, color = "red", border_right = TRUE)
min_orders <- min(orders_sort)
max_orders <- max(orders_sort)
n_bins <- 5
bin_width <- (max_orders - min_orders) / n_bins
library(dplyr)
orders_df <- data_frame(orders = orders_sort)
orders_bin <- orders_df %>% 
  mutate(category=cut(orders, breaks=c(-Inf, min(orders) + 1:4*round((max(orders) - min(orders)),0) / 5, Inf), labels=c("very low","low","medium","high", "very high"))) %>% 
  group_by(category) %>% 
  add_tally() %>% 
  group_by(category) %>% 
  summarize ( frequency = mean(n) ) %>% 
  mutate(begin = min(orders) + 0:4*(max(orders)-min(orders))/length(category), end = min(orders) + 1:5*(max(orders)-min(orders))/length(category)) %>% 
  mutate(midpoint = (begin + end) / 2) %>% select(category, begin, end, midpoint, frequency)
## 
```
The data has also been re-indexed as shown in the top row of the table.

2. Choose the number of bins (groups, intervals). Let's choose 5 bins. Usually bins are an odd, not an even, number. They typically range from 3 to 9 in practice.

3. Calculate the bin width. We let $w$ be the width, $n$ the number of bins and $x$ the sorted orders in the formula below.

$$
w = \frac{max(x)-min(x)}{n} = \frac{`r max_orders` - `r min_orders`}{`r n_bins`} = `r bin_width`
$$
Already we are binning to describe our data. We have sorted it from lowest value $min(x)$ to the highest value $max(x)$. We are beginning to group the data into $n=`r n_bins`$.

4. In a vertical table arrange the bins from lowest interval to highest. Initially use 5 columns as in the table below. The intervals will each have a beginning and ending value such that groups of orders will lie in non-overlapping intervals of bin-width.

$$
begin \leq orders < end
$$
We do not want to ever double count the number of orders in a bin, thus the $<$ relation for the end. The first $begin$ value is the minimum order $`r min_orders`$. The bin-width is $`r bin_width`$. The ending value of the first interval is then 

$$
end = begin + width = `r min_orders` + `r bin_width` = `r min_orders + bin_width` 
$$

Thus our first interval looks like this

$$
`r min_orders` \leq orders < `r min_orders + bin_width`
$$

The interval or class midpoint is the arithemetic average of the interval from beginning to end. So for the first group

$$
midpoint = \frac{begin+end}{2} = \frac{`r min_orders` + `r min_orders + bin_width`}{2} = `r (min_orders + min_orders + bin_width)/2`
$$

Because we have sorted our data, it is a simple exercise to count the number of orders in this first bin just by examining the sorted series.

<br>
How many are there?
<br>
<button onclick="showText('mydiv10')">show / hide</button>
<div id="mydiv10" style="display:none;">
<br>

Just 1!

</div>

<br>

The relationships in the last interval are very important to remember. From the very low to the high intervals the relationships are

$$
begin \leq orders < end
$$

But to obery the rubric that we must use all of our data (and also remember not to every double count), the last, the very high interval has these relationships.

$$
begin \leq orders \leq max(orders)
$$

We must always remember to use the $\leq$ relationship to include the $max(x)$ of our data series.

Let's build out the table  with 5 columns: category, begin, end, midpoint, frequency. Try this on paper.

<br>
<button onclick="showText('mydiv11')">show / hide</button>
<div id="mydiv11" style="display:none;">
<br>

```{r table-1}
table_data = orders_bin
table_data %>% 
  kable(align = "c") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>% 
  column_spec(1, bold = T, color = "red", border_right = TRUE)
```
<br>
</div>

## Review these scenarios

1. How can a production manager in a start-up summarize and describe to the board of directors the results of testing the life of a sample of 100 revolutionary light-weight organic batteries?

<br>
<button onclick="showText('myDIV11')">show / hide</button>
<div id="myDIV11" style="display:none;">

There is too much raw data to show the board. The manager wants to summarize the results by displaying the average life and range of lives of the batteries. The manager might also divide the sample into low, medium, and high durations to illustrate the distribution of lives in the sample. Each of these is an example of descriptive statistics.

</div>
<br>

2. Why would the manager want to infer that the sample is representative of the whole output of batteries?

<br>
<button onclick="showText('myDIV12')">show / hide</button>
<div id="myDIV12" style="display:none;">

The manager can't test the life of all of the batteries -- that would destroy the output of the venture! Instead the manager might want to test the hypothesis that the average battery life is greater than, say, 1,000 hours, with a precision of 5% that the manager is wrong about this hypothesis.

</div>
<br>

3. Identify the population and sample in this situation.

<br>
<button onclick="showText('myDIV13')">show / hide</button>
<div id="myDIV13" style="display:none;">

The manager takes a sample of 100 batteries from a population that includes all of the venture's output. The manager would like to describe the sample in a way that is probably representative of the output of the venture.

</div>
<br>

4. The manager is putting together a testing database.

a. What kind of data is the average life of the batteries? 

<br>
<button onclick="showText('myDIV14')">show / hide</button>
<div id="myDIV14" style="display:none;">

Since life is greater than or equal to zero, life data would numeric. Because I can always find a life value between any two consecutive life values, the data is continuous.

</div>
<br>

b. What about the number of batteries in each location?

<br>
<button onclick="showText('myDIV15')">show / hide</button>
<div id="myDIV15" style="display:none;">

All counts are integers and thus discrete.

</div>
<br>

c. What if the batteries were produced in various locations. What kind of data might the locations be? 

<br>
<button onclick="showText('myDIV16')">show / hide</button>
<div id="myDIV16" style="display:none;">

Location is a classification and thus nominal.

</div>
<br>

d. The manager asks experts to rate the reliability of the batteries. What kind of data are the ratings?

<br>
<button onclick="showText('myDIV17')">show / hide</button>
<div id="myDIV17" style="display:none;">

Ratings rank in some order from bad to good, or unlikely to likely, or some other ordered range, and thus ordinal.

</div>

## Try this

Apply the data hierarchy to the county health outcome rankings and data base for states in the US. Find the types of data and apply the canons of empirical method to your findings. Choose Other measures > socio-economic factors for your investigation. [You can navigate to this site for an interactive view.](https://www.countyhealthrankings.org/app/new-york/2019/measure/outcomes/1/data)

1. [Download this report](https://www.countyhealthrankings.org/sites/default/files/state/downloads/CHR2019_NY.pdf)

2. Identify at least two examples each of nominal, ordinal, interval, and ratio variables.

3. Upon reading this report, are appropriate data types reach valid conclusions? For example, if a conclusion reached indicates that one segment of the sampled population is twice as large as the sampled population of another segment, are ratio measurement scales used?
