---
title: "Workbook: how probable?"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(bsselectR))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(stringr))
```

<script>
function showText(y) {
    var x = document.getElementById(y);
    if (x.style.display === "none") {
        x.style.display = "block";
    } else {
        x.style.display = "none";
    }
}
</script>

##  Age and gender of U.S. physicians in a recent year, as reported by the American Medical Association.

```{r}
data <- read.csv("data/qq-4-37.csv")
colnames(data) <- c("factor",	"< 35",  "35-44",	"45-54",	"55-65",	"> 65",	"gender")
data %>%
  kable("html", escape = F) %>% 
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = F)
```


a. What is the probability that one randomly selected physician is 35-44 years old?

<button onclick="showText('myDIV1')">show / hide</button>
<div id="myDIV1" style="display:none;">

.28

</div>

b. What is the probability that one randomly selected physician is both a woman and 45-54 years old?

<button onclick="showText('myDIV2')">show / hide</button>
<div id="myDIV2" style="display:none;">

.04

</div>

c. What is the probability that one randomly selected physician is a man or is 35-44 years old?

<button onclick="showText('myDIV3')">show / hide</button>
<div id="myDIV3" style="display:none;">

.86

</div>

d. What is the probability that one randomly selected physician is less than 35 years old or 55-64 years old?

<button onclick="showText('myDIV4')">show / hide</button>
<div id="myDIV4" style="display:none;">

.32

</div>

e. What is the probability that one randomly selected physician is a woman if she is 45-54 years old?

<button onclick="showText('myDIV5')">show / hide</button>
<div id="myDIV5" style="display:none;">

.1739

</div>

f. What is the probability that a randomly selected physician is neither a woman nor 55-64 years old?

<button onclick="showText('myDIV6')">show / hide</button>
<div id="myDIV6" style="display:none;">

.66

</div>